\resumenIngles{

The main objective of this thesis is verify if the efficiency of recommendation systems (RS) improves when the communities are pre calculated. The output of RS are recommendations presented to an active user and intends to guide his navigation on contents of their interest. Recommendations are based on different aspects, one of those aspects are the communities to the user belongs. This thesis has been supported on a framework of collaborative events called 3-Ontology, implementing an API called RBox 2.0, where the communities are one of the essentials dimensions.

The result of the interactions between users and Web 2.0 applications is the apparition of many communities. The Twitter followers of an user forms a community as well as the buyers of a same book on Amazon. The RS  uses communities and typically different detection strategies are needed to identify them. However, those strategies are not backed on a mechanism that entirely supports the recommendation process, and the detection results must be reprocessed in every recommendation. The fact of counting with a framework that standardizes the representation of every interaction produced in a social media environment permits that the community detection may be realized in a standard way. As a plugin to RBox, we present a mechanism of community abstraction, representation, detection and persistence (community cache), that permits handling a backup of the communities that have been already detected in past recommendations.

In order to validate the efficacy of the proposed model, we have developed a recommendation system, that uses the model, and performs recommendations based on different community detection strategies, based on literature. The testing scenarios show that the average recommendation time is seven percent lower when the recommendations are performed with the proposed model. As a contribution to the RS area, we present an abstraction layer for the Rbox framework, that permits to detect, persist and use efficiently in the recommendation, the information related to different emerging communities.

\vspace*{0.5cm}

\KeywordsEN{community detection, recommendation systems, Web 2.0, 3-Ontology, community cache, social networking}
}
